---
title: "COVID-19 Trends in Rwanda (Apr 2020 - Jul 2021)"
author: "Sarah Agatoni"
output:pdf_document: default
---

<br> 

## 0: Summary

This project uses data from the Rwandan Ministry of Health Twitter account to: 
<br>

- explore the spread of COVID-19 in Rwanda between 17 April 2020 to 23 July 2021

- explore how COVID-19 impacted various demographics during this period


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, results='hide'}
setwd("~/Documents/Data Projects/")
```
<br>

## 1: Data Loading & Prep
<br>

**loading packages**
```{r packages, results='hide', message=FALSE}
if (!require(pacman)) install.packages(pacman); library(pacman)  
p_load(broom, choroplethrAdmin1, data.table, dplyr, ggplot2, ggpubr, 
       ggthemr, janitor, lemon, lubridate, maps, rgdal, sp)
```
<br>


**reading data in csv format**
```{r data}
#data manually obtained from Rwanda Ministry of Health Twitter
data <- data.table::fread("covid_rw.csv")
demo <- data.table::fread("covid_demo.csv")
```
<br>


**clean-up**
```{r clean}
data$Date <- dmy(data$Date)
demo$Date <- dmy(demo$Date)
data <- janitor::clean_names(data)
demo <- janitor::clean_names(demo)
```
<br>


## 2: Data Viz: Positivity Trend 
<br>


**Data Prep**
```{r transform_1}
#calculating positivity rate as new daily cases per tests administered
data <- data %>%
  mutate(pos_rate = new_cases/tests_today)
```
<br>


**Line Graph: Positivity Trend**
```{r plot_1, fig.width=10, fig.height=6}
ggthemr("fresh")
plot_1 <- ggplot(data = data, mapping = aes(x = date, y = pos_rate)) + 
  geom_line(col = 'light grey') + 
  geom_smooth(col = '#B8390E', se = FALSE) + 
  scale_y_continuous("Positivity Rate", labels = scales::percent) +
  scale_x_date("Date", date_breaks = "1 month", date_labels = '%b') +
  no_gridlines() +
  theme(plot.margin = unit(c(1,1,1,1), "cm"))
plot_1
```
<br> 
<br> 


## 3: Data Viz: Deaths by Age & Gender
<br>


**Data Prep**
```{r by_age}
# grouping death counts by age and gender 
by_age <- na.omit(demo) %>%
  
  mutate(gender = case_when(gender == "F" ~ "Female", gender == "M" ~"Male"),
         agegroup = case_when(age <= 19 ~ '19 & Under',
                              age >= 20  & age <= 29 ~ '20s',
                              age >= 30  & age <= 39 ~ '30s',
                              age >= 40  & age <= 49 ~ '40s',
                              age >= 50  & age <= 59 ~ '50s',
                              age >= 60  & age <= 69 ~ '60s',
                              age >= 70  & age <= 79 ~ '70s',
                              age >= 80  & age <= 89 ~ '80s',
                              age >= 90  & age <= 99 ~ '90s',
                              age >= 100 ~ '100 & Over',
                              is.na(age) == TRUE ~ "Unclear"),
         agegroup = factor(agegroup, 
                           levels = c("Unclear", "19 & Under", "20s", "30s", "40s",
                                      "50s", "60s", "70s", "80s", "90s", "100 & Over"), 
                           ordered = TRUE)) %>%
  
 count(agegroup, gender, name = "fatalities")
  
```
<br>


**Population Pyramid: Deaths by Age & Gender**
```{r pop_plot, fig.width=10, fig.height=6}
ggthemr('dust')
plot_2 <- ggplot(
  data = by_age, 
  mapping = aes(x = ifelse(test = gender == "Female", yes = -fatalities, no = fatalities), y = agegroup, fill = gender)) + 
  geom_col() + 
  lemon::scale_x_symmetric(labels = abs) + 
  labs(x = "Number of People", y = "Age", fill = "Gender") + 
  theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  no_gridlines() 

plot_2
```
<br> 
<br> 


## 4: Data Viz: Deaths by District
<br>


**Data Prep**
```{r by_district, message=FALSE}

dist <- na.omit(demo) %>%
  group_by(district) %>%
  count(name = "deaths")

#adding Ngoma district; missing in original data file due to zero deaths
dist <- as.data.frame(dist)
by_dist <- rbind(dist, c("Ngoma",0))
by_dist$deaths <- as.integer(by_dist$deaths)


#normalizing death counts by district i.e. calculating deaths per population
#population counts obtained from PEPFAR  
#https://www2.census.gov/programs-surveys/international-programs/tables/time-series/pepfar/rwanda.xlsx
by_dist <- by_dist %>%
 mutate(population = case_when(
    district == "Kigali"~ 1577326,
    district == "Nyarugenge" ~ 338862,
    district == "Gasabo" ~ 795658,
    district == "Kicukiro" ~ 442806,
    district == "Nyanza" ~ 410717,
    district == "Gisagara" ~ 361573,
    district == "Nyaruguru" ~ 341196,
    district == "Huye" ~ 369480,
    district == "Nyamagabe" ~ 381677,
    district == "Ruhango" ~ 374582,
    district == "Muhanga" ~ 331126,
    district == "Kamonyi" ~ 401146,
    district == "Karongi" ~ 361647,
    district == "Rutsiro" ~ 361121,
    district == "Rubavu" ~ 526725,
    district == "Nyabihu" ~ 301035,
    district == "Ngororero" ~ 359828,
    district == "Rusizi" ~ 443967,
    district == "Nyamasheke" ~ 409888,
    district == "Rulindo" ~ 303379,
    district == "Gakenke" ~ 331763,
    district == "Musanze" ~ 423647,
    district == "Burera" ~ 338043,
    district == "Gicumbi" ~ 432700,
    district == "Rwamagana" ~ 393112,
    district == "Nyagatare" ~ 767710,
    district == "Gatsibo" ~ 607320,
    district == "Kayonza" ~ 489568,
    district == "Kirehe" ~ 440339,
    district == "Ngoma" ~ 424047,
    district == "Bugesera" ~ 444823),
    
    norm_fat = (deaths/population) * 100000
    )

#reading map data as shape files
#obtained from Stanford's Libraries: https://earthworks.stanford.edu/catalog/stanford-qy869sx9298)

my_spdf <- readOGR(dsn = ("~/Documents/Data Projects/rwshapefiles/"))


#converting shape files to tidy data  
#recoding "gasabo", "kicukiro" and "nyarugenge" as "kigali"; covid-19 data for Kigali isn't stratified by district 
spdf_tidy <- tidy(my_spdf, region = "NAME_2") %>%
  mutate(group = as.character(group),
         region = case_when(
           grepl("Gasabo|Kicukiro|Nyarugenge", group, ignore.case = TRUE) ~ "Kigali",
           TRUE ~ id
    ))
  
#relating map data with covid-19 data 
choro_df <- left_join(spdf_tidy, by_dist, by = c("region" = "district"))

#adding district labels
dist_names <- spdf_tidy %>%
  group_by(region) %>%
  summarise(long = mean(long), lat = mean(lat))

```
<br>


**Choropleth Map: Deaths by District**
```{r fig.width=10, fig.height=8}
plot_4 <- ggplot() +
  
  #gradient map
  geom_polygon(data = choro_df, 
               mapping = aes(x = long, y = lat, group = group, fill = norm_fat), color = "dark grey", size = 0.05) +
  theme_void() +
  
  #color scheme with horizontal legend at the bottom
  scale_fill_gradient(
    low ="#fee7e5",
    high = "#fc1100",
    guide = guide_colorbar(title = "deaths per 100,000 people",
                           direction = "horizontal",
                           barheight = unit(2, units = "mm"),
                           barwidth = unit(42, units = "mm"),
                           title.position = 'top',
                           title.theme = element_text(size = 10))) +
  
  #legend position & plot margins
  theme(legend.position = "bottom", 
        plot.margin = unit(c(1,1,1,1), "cm")) +
  
  #adding district names 
  geom_text(data = dist_names, 
           mapping =  aes(x = long, y = lat, label = region), 
           size = 3, hjust = "center") +
  
  #fixing aspect ratio
  coord_fixed(0.9) 

plot_4
```




